Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-09 conclusions.tex
lloid_paper: polishing pass post-Steve
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@986 ccdd05d5-56c1-461a-bf98-85f6876523b4

We have demonstrated a computationally feasible procedure for the rapid and
even preemptive detection of \GW{}s emitted during the coalescence
of neutron stars and stellar-mass black holes. Our method is as fast as
standard \fft{} convolutions but allows for zero latency, sample-in-sample-out
operation.  Our search targets are expected to produce prompt electromagnetic
signals and may be the progenitors of some short hard gamma-ray bursts.  Rapid
alerts to the broader astronomical community will improve the chances of
detecting an electromagnetic counterpart in bands from gamma-rays down to
radio.  We anticipate requiring $\sim100$ modern multi-core computers to
analyze a four-detector network of \GW{} data for binary neutron
stars and stellar-mass black holes.  This is well within the current computing
capabilities of the \LIGO{} Data Grid~\cite{LDG}. In the future, this could be
improved upon further by conditionally reconstructing the \SNR{} time-series
only during times when a composite detection statistic crosses a
threshold~\cite{svd-compdetstat}.

>>>>

We have demonstrated a computationally feasible procedure for the rapid and
even preemptive detection of \GW{}s emitted during the coalescence
of neutron stars and stellar-mass black holes. Our method is as fast as
standard \fft{} convolutions but allows for zero latency, sample-in-sample-out
operation.  Our search targets are expected to produce prompt electromagnetic
signals and may be the progenitors of some short hard gamma-ray bursts.  Rapid
alerts to the broader astronomical community will improve the chances of
detecting an electromagnetic counterpart in bands from gamma-rays down to
radio.  We anticipate requiring $\sim$100 modern multi-core computers to
analyze a four-detector network of \GW{} data for binary neutron
stars and stellar-mass black holes.  This is well within the current computing
capabilities of the \LIGO{} Data Grid~\cite{LDG}. In the future, this could be
improved upon further by conditionally reconstructing the \SNR{} time-series
only during times when a composite detection statistic crosses a
threshold~\cite{svd-compdetstat}.

### 2011-06-09 introduction.tex
lloid_paper: polishing pass post-Steve
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@986 ccdd05d5-56c1-461a-bf98-85f6876523b4

The coalescence of compact binary systems consisting of neutron stars
 (\textsc{ns}) and/or black holes (\textsc{bh}) is presently the best
understood and most actively pursued target source of gravitational
radiation for ground-based gravitational wave (\GW{}) detectors.
As a compact binary system loses energy to \GW{}s, its
orbital separation decays, leading to a run-away inspiral with the \GW{}
amplitude and frequency increasing until the system eventually merges. At the
merger, our understanding of the system's behavior is incomplete. It is thought
that if a neutron star is involved, it may become tidally disrupted near
the merger and fuel a bright electromagnetic counterpart ~\cite{shibata:2007}.
This counterpart is expected to be rich in information about the source, but 
extracting that information requires rapid identification of candidate
gravitational wave events.

Though compact binary coalescence (\CBC{}) events may be responsible for several
 classes of electromagnetic transients, \CBC{}s are in particular believed to be a
mechanism for short gamma-ray bursts (short \GRB{}s)~\cite{Lee:2005, nakar07}.
In this scenario, prompt electromagnetic emission arises as shells of
 relativistically out-flowing matter collide in the inner shock.
The same inner shocks, or potentially reverse shocks, can produce a bright
 accompanying optical flash~\cite{Sari99}. Prompt emission is a probe into
 the extreme initial conditions of the outflow, in contrast with afterglows,
which arise in the external shock with the local medium and are relatively
insensitive to initial conditions.

Optical flashes typically peak in brightness within tens of seconds and
consequently have only been observed for a handful of long
\GRB{}s~\cite{2011CRPhy..12..255A} by telescopes with extremely rapid response
or, in the case of \textsc{grb 080319b}, by pure serendipity, where several
telescopes were observing a previous \GRB{} in the same
field~\cite{2008Natur.455..183R}. Short \GRB{}s, on the other hand, typically
fade too quickly to even catch the tails of the afterglows. Rapid
\GW{} transient alerts could enable the first observation of
prompt optical flashes and the rise of afterglows from short \GRB{}s, and, for
an exceptional event, perhaps even a glimpse of a
precursor~\cite{0004-637X-723-2-1711}.

Observation of an electromagnetic counterpart to a candidate \GW{} event would
vastly boost confidence in the \GW{} detection and provide the best possible
estimation of the event's positon in the sky localization. With such detailed
information, it will be possible to determine the source's host
galaxy and therefore measure the source's redshift. A coincident \GW{}
observation of this event will also provide a luminosity distance measurement.
Together, these two quantities immediately lead to measurements of the Hubble
constant of unprecedented accuracy ~\cite{2010ApJ...725..496N}.

In the era of first-generation detectors, the \GW{} community initiated a project
to send alerts when potential \GW{} transients were observed in
order to trigger followup observations by optical telescopes.  The best
demonstrated latencies were 30--60 minutes~\cite{HugheyGWPAW2011}, which was an
important achievement, but far too late to catch prompt electromagnetic
emission. In order to extract the maximal science from these events, we have 
the ambition of reporting \GW{} candidates not minutes \emph{after} the merger, 
as has already been accomplished, but seconds \emph{before}.

We show below that by examining the signal to noise ratio (\SNR{}) stream for
threshold crossings before the \GW{} signal leaves the detection band, it is
possible to trade some \SNR{} and sky localization
accuracy~\cite{Fairhurst2009} for latency.  Figure \ref{fig:earlywarning} shows
projected early trigger rates for \textsc{ns}--\textsc{ns} binaries in Advanced
\LIGO\ assuming the event rate predictions in~\cite{Abadie:2010p10836}.  The
most likely estimates indicate that 10 sources could be audible (meaning
producing an \SNR\ above 8) 10 seconds or more prior to the merger during one
year of observation.  This number is highly uncertain.  In fact as many as 100
events detected 10 seconds early is plausible. It is also plausible that no
sources will be detected at all.  The gray bands give the pessimistic to
optimistic interval defined in ~\cite{Abadie:2010p10836}. 
%
\begin{figure}
\begin{center}
\includegraphics{figures/snr_in_time.pdf}
\caption{\label{fig:earlywarning} Expected number of \textsc{ns}--\textsc{ns}
sources that will be detectable $t$ seconds before coalescence.  The heavy
solid line is the most likely yearly rate estimate $\dot N_{\mathrm{re}}$
during Advanced \LIGO.  The shaded region represents the confidence interval
$[\dot N_{\mathrm{low}}, \dot N_{\mathrm{high}}]$ described in
\cite{Abadie:2010p10836}.  Advanced \LIGO\ noise projections are described in
\cite{ALIGONoise}.  Assuming that an \SNR\ of 8 is sufficient for detection and
that we observe $\dot N_{\mathrm{re}} = 40$ events~yr$^{-1}$ with a detector in
the `zero detuning, high power' configuration, about 10~yr$^{-1}$ will be
detectable within 10 seconds before merger and $\sim1$~yr$^{-1}$ will be
detectable within 100 seconds of merger.} 
\end{center}
\end{figure}

In October 2010, \LIGO{} completed its sixth science run
(S6) and Virgo completed its third science run (VSR3).  While both
\LIGO{} detectors and Virgo were operating, several all-sky detection
pipelines operated in a low-latency configuration to send astronomical alerts,
namely \textsc{mbta}, Coherent WaveBurst, and
Omega~\cite{HugheyGWPAW2011, S6lowlatency}.
\editorial{Get references for these low-latency pipelines.} The S6 analyses
achieved latencies of 30--60 minutes, which were dominated by a human vetting
process. Candidates were sent for electromagnetic followup to several
telescopes; Swift, \textsc{rotse}, \textsc{tarot}, and Zadko~\cite{kanner2008,
HugheyGWPAW2011} took images of likely sky locations.  \textsc{mbta} achieved
the best \GW{} trigger-generation latencies of 2--5 minutes.  We
assume that in the advanced detector era the vetting process will be automated,
so current \GW{} search methodology and telescope actuation would
dominate latency.

Advance detection of compact binary coalescences (\CBC{}s) will require striking a balance between latency
and throughput.  \CBC{} searches consist of banks of matched filters, or
cross-correlations between the data stream and a bank of nominal ``template''
signals.  There are many different implementations of matched filters, but most
have high throughput at the cost of high latency, or low latency at the cost of
low throughput.  The former are epitomized by the overlap-save algorithm
\cite{numerical-recipes-chapter-13} for frequency domain (\FD) convolution,
currently the preferred method in \GW{}
searches.  The most obvious example of the latter is the time domain
(\TD) convolution, which has no algorithmic latency.  However, its
computational complexity is quadratic in the length of the templates, so it is
prohibitively expensive for long templates.

>>>>

The coalescence of compact binary systems consisting of neutron stars
 (\textsc{ns}) and/or black holes (\textsc{bh}) is presently the best
understood and most actively pursued target source of gravitational
radiation for ground-based gravitational-wave (\GW{}) detectors.
As a compact binary system loses energy to \GW{}s, its
orbital separation decays, leading to a run-away inspiral with the \GW{}
amplitude and frequency increasing until the system eventually merges.  It is
thought that if a \textsc{ns} is involved, it may become tidally disrupted near
the merger and fuel a bright electromagnetic (\EM{})
counterpart~\cite{shibata:2007}. Observation of an \EM{} counterpart to a
candidate \GW{} event will vastly boost confidence in the \GW{} detection and
will often provide sufficient sky localization to determine
the source's host galaxy and therefore measure the source's redshift. The
\GW{} observation of this event will also provide a luminosity
distance measurement. Together, these two quantities immediately lead to
measurements of the Hubble constant of unprecedented
accuracy~\cite{2010ApJ...725..496N}. The most efficient means of obtaining
such joint observations is through the rapid identification of candidate \GW{}
events.

Though compact binary coalescence (\CBC{}) events may be responsible for several
classes of \EM{} transients, \CBC{}s are in particular believed to be a
mechanism for short gamma-ray bursts (short \GRB{}s)~\cite{Lee:2005, nakar07}.
In this scenario, prompt \EM{} emission arises as shells of
relativistically out-flowing matter collide in the inner shock.
The same inner shocks, or potentially reverse shocks, can produce a bright
accompanying optical flash~\cite{Sari99}. Prompt emission is a probe into
the extreme initial conditions of the outflow, in contrast with afterglows,
which arise in the external shock with the local medium and are relatively
insensitive to initial conditions.

Optical flashes have only been observed for a handful of long
\GRB{}s~\cite{2011CRPhy..12..255A} by telescopes with extremely rapid response
or, in the case of \textsc{grb 080319b}, by pure serendipity, where several
telescopes were observing a previous \GRB{} in the same
field~\cite{2008Natur.455..183R}. The observed optical flashes peaked
within tens of seconds and decayed quickly.
Short \GRB{}s, on the other hand, typically
fade too quickly to even catch the tails of the afterglows in any band. Rapid
\GW{} transient alerts could enable the first observation of
prompt optical flashes and the rise of afterglows from short \GRB{}s, and, for
an exceptional event, perhaps even a glimpse of a
precursor~\cite{0004-637X-723-2-1711}.

In the era of first-generation detectors, the \GW{} community initiated a
project to send alerts when potential \GW{} transients were observed in
order to trigger followup observations by optical telescopes.  The best
demonstrated latencies were
$\mathcal{O}(1\textrm{ hour})$~\cite{HugheyGWPAW2011}, which was an important
achievement, but far too late to catch prompt \EM{}
emission. In order to extract the maximum science from Advanced \LIGO{}
\CBC{} events, we have the ambition of reporting \GW{} candidates not minutes
\emph{after} the merger, as has already been accomplished, but seconds
\emph{before}.

We show that by examining the signal to noise ratio (\SNR{}) stream for
threshold crossings before the \GW{} signal leaves the detection band, it is
possible to trade some \SNR{} and sky localization
accuracy for negative latency.  Figure~\ref{fig:earlywarning} shows
projected early trigger rates for \textsc{ns}--\textsc{ns} binaries in Advanced
\LIGO\ assuming the event rate predictions in~\cite{Abadie:2010p10836}.  The
most likely estimates indicate that $\sim$10 sources will produce an \SNR\
above 8 in the detector 10 seconds or more prior to the merger during one
year of observation, though the uncertainty spans orders of magnitude.
The gray bands give the pessimistic to optimistic interval defined
in~\cite{Abadie:2010p10836}.
%
\begin{figure}
\begin{center}
\includegraphics{figures/snr_in_time.pdf}
\caption{\label{fig:earlywarning} Expected number of \textsc{ns}--\textsc{ns}
sources that will be detectable $t$ seconds before coalescence.  The heavy
solid line is the most likely yearly rate estimate $\dot N_{\mathrm{re}}$
during Advanced \LIGO.  The shaded region represents the confidence interval
$[\dot N_{\mathrm{low}}, \dot N_{\mathrm{high}}]$ described in
\cite{Abadie:2010p10836}.  Advanced \LIGO\ noise projections are described in
\cite{ALIGONoise}.  Assuming that an \SNR\ of 8 is sufficient for detection and
that we observe $\dot N_{\mathrm{re}} = 40$ events~yr$^{-1}$ with a detector in
the `zero detuning, high power' configuration, $\sim$10~yr$^{-1}$ will be
detectable within 10 seconds before merger and $\sim$1~yr$^{-1}$ will be
detectable within 100 seconds of merger.} 
\end{center}
\end{figure}

In October 2010, \LIGO{} completed its sixth science run
(S6) and Virgo completed its third science run (VSR3).  While both
\LIGO{} detectors and Virgo were operating, several all-sky detection
pipelines operated in a low-latency configuration to send astronomical alerts,
namely \textsc{mbta}, Coherent WaveBurst, and
Omega~\cite{HugheyGWPAW2011, S6lowlatency}.
\editorial{Get references for these low-latency pipelines.} The S6 analyses
achieved latencies of 30--60 minutes, which were dominated by a human vetting
process. Candidates were sent for \EM{} followup to several
telescopes; Swift, \textsc{rotse}, \textsc{tarot}, and Zadko~\cite{kanner2008,
HugheyGWPAW2011} took images of likely sky locations.  \textsc{mbta} achieved
the best \GW{} trigger-generation latencies of 2--5 minutes.  We
assume that in the advanced detector era the vetting process will be automated,
so current \GW{} search methodology and telescope actuation would
dominate latency.

Realizing advance detection of compact binary coalescences (\CBC{}s) will
require striking a balance between latency and throughput. \CBC{} searches
consist of banks of matched filters, or
cross-correlations between the data stream and a bank of nominal ``template''
signals.  There are many different implementations of matched filters, but most
have high throughput at the cost of high latency, or low latency at the cost of
low throughput.  The former are epitomized by the overlap-save algorithm
\cite{numerical-recipes-chapter-13} for frequency domain (\FD) convolution,
currently the preferred method in \GW{}
searches.  The most obvious example of the latter is the time domain
(\TD) convolution, which has no algorithmic latency.  However, its
computational complexity is quadratic in the length of the templates, so it is
prohibitively expensive for long templates.

### 2011-06-09 results.tex
lloid_paper: polishing pass post-Steve
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@986 ccdd05d5-56c1-461a-bf98-85f6876523b4

Table~\ref{table:flops} shows that the \lloid\ method requires $\sim 10^8$
\flops\ to cover a sub-bank comprising $\sim 10^2$ out of the total $\sim 10^5$
mass pairs.  Given that modern (ca. 2011) workstations can sustain computation
rates up to $\sim 10^{10}$ \flops{}, and assuming that other regions of the
parameter space have similar computational scaling, an entire search could be
implemented with $\gtrsim 10$ machines.  The lengths of templates does vary
over the parameter space; lower-mass templates are longer and would require
more computation to analyze while higher-mass templates are shorter and would
require less. However, we consider the estimates based on this sub-bank to be a
reasonable representation.

By comparison, using the \TD\ method to achieve the same latency costs $\sim
10^{13}$ \flops\ for this particular sub-bank, and so it would require $\sim
10^6$ present-day machines to search the full parameter space.  Presently, the
\LIGO{} Data Grid consists of only $\sim 10^4$ machines.

>>>>

Table~\ref{table:flops} shows that the \lloid\ method requires $\sim$$10^8$
\flops\ to cover a sub-bank comprising $\sim$$10^2$ out of the total $\sim$$10^5$
mass pairs.  Given that modern (ca. 2011) workstations can sustain computation
rates up to $\sim$$10^{10}$ \flops{}, and assuming that other regions of the
parameter space have similar computational scaling, an entire search could be
implemented with $\gtrsim$$10$ machines.  The lengths of templates does vary
over the parameter space; lower-mass templates are longer and would require
more computation to analyze while higher-mass templates are shorter and would
require less. However, we consider the estimates based on this sub-bank to be a
reasonable representation.

By comparison, using the \TD\ method to achieve the same latency costs
$\sim$$10^{13}$ \flops\ for this particular sub-bank, and so it would require
$\sim$$10^6$ present-day machines to search the full parameter space.
Presently, the \LIGO{} Data Grid consists of only $\sim$$10^4$ machines.
